---
title: "Tidyverse Essentials"
author: "Kaymal"
output: 
  github_document:
    toc: yes
    
---

- Data Wrangling/Transformation with `dplyr`
- Data Visualization with `ggplot2`


```{r}
# Load the necessary packages
library(gapminder)
library(dplyr)
# vignette("dplyr")
```


```{r}
# Print the head of the  gapminder dataset
head(gapminder, n = 3)

# Print the tail of the  gapminder dataset
tail(gapminder, n = 3)
```

```{r}
summary(gapminder)
```


## Data Wrangling with `dplyr`

I'll use `dplyr` to apply transformation such as filtering, sorting, and summarizing to data. `dplyr` is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges:

- `filter()` picks/selects cases based on their values.
- `arrange()` changes the ordering of the rows.
- `mutate()` adds new variables that are functions of existing variables.
- `select()` picks variables based on their names.
- `summarize()` reduces multiple values down to a single summary.

These all combine naturally with `group_by()` which allows to perform any operation “by group”. 

[dplyr](https://dplyr.tidyverse.org)

### The `filter` Verb

`filter()` allows us to select a subset of rows in a data frame. 

```{r}
# Filter the gapminder dataset for the country Turkey
gapminder %>%
  filter(country == "Turkey")
```

```{r}
# Filter for Turkey in 2007
gapminder %>% 
  filter(country == "Turkey", year == 2007)
```

### The `arrange` Verb

`arrange()` works similarly to `filter()` except that instead of filtering or selecting rows, it reorders them.

```{r}
# Sort in ascending order of lifeExp
gapminder %>%
  arrange(lifeExp)

  
# Sort in descending order of lifeExp
gapminder %>% 
  arrange(desc(lifeExp))
```


```{r}
# Filter for the year 2008, then arrange in descending order of population
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(pop))
```

### The `select` Verb

`select()` allows us to rapidly zoom in on a useful subset using operations that usually only work on numeric variable positions.

```{r}
gapminder %>%
  select(country, pop)
```

We can also return a “subset” of columns from a data frame by listing the name of each column to drop.

```{r}
gapminder %>%
  select(-c(continent, pop, gdpPercap))
```


### The `mutate` Verb

Besides selecting sets of existing columns, it’s often useful to add new columns/variables that are functions of existing columns. 

```{r}
# Filter, mutate, and arrange the gapminder dataset
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>% 
  filter(year == 2007) %>%   
  arrange(desc(gdp))
```

### The `summarize` Verb

`summarize` helps us compute summary statistics for the data (e.g `sum()`, `max()`, `mean()`).

```{r}
# Summarize to find the mean life expectancy
gapminder %>%
  summarize(meanLifeExp = mean(lifeExp))
```

```{r}
# Filter for 2007 then summarize the mean and median life expectancy
gapminder %>%
  filter(year == 2007) %>%
  summarize(meanLifeExp = mean(lifeExp),
            medianLifeExp = median(lifeExp))
```


### Grouping Data with `group_by`Verb

```{r}
# Find the mean life expectancy and total population in each year
gapminder %>%
  group_by(year) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(as.numeric(pop)))
```

Both mean and median life expectancy are going up over time.

```{r}
# Find mean and median life expectancy in each year/continent combination
gapminder %>%
  group_by(year, continent) %>%
  summarize(meanLifeExp = mean(lifeExp),
            medianLifeExp = median(lifeExp))
```

## Visualization with `ggplot2`

`ggplot2` is a system for 'declaratively' creating graphics, based on "The Grammar of Graphics". 

```{r}
library(ggplot2)
```

```{r}
# Create gapminder_2007
gapminder_2007 <- gapminder %>%
  filter(year == 2007)
```

```{r}
# Plot gdpPercap vs lifeExp for the year 2007
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

The problem with this plot is that, most of the countries are in the leftmost part of the x-axis. Let's plot the graph with population on log scale. 

### Log Transformation

```{r}
# Plot log transformed gdpPercap vs lifeExp for the year 2007
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  scale_x_log10()
```

The relationship between GDP per capita and life expetancy looks more **linear** on this scale. Additionally, it is now easier to distinguish the countries at the lower end of the spectrum.

```{r}
# Plot histogram of gdpPercap
ggplot(gapminder_2007, aes(x = gdpPercap)) +
  geom_histogram(bins = 10)

# Plot histogram of log(gdpPercap)
ggplot(gapminder_2007, aes(x = gdpPercap)) +
  geom_histogram(bins = 10) +
  scale_x_log10()
```

On a log scale, the distribution of `gdpPercap` is approximately symmetrical.


```{r}
# Plot pop vs lifeExp
ggplot(gapminder_2007, aes(x = pop, y = lifeExp)) +
  geom_point() 
```

It is a good idea to put the x-axis on a log scale to better undrestand the relationship between population and the life expectancy.

```{r}
# Plot pop vs lifeExp with the x-axis on a log scale
ggplot(gapminder_2007, aes(x = pop, y = lifeExp)) +
  geom_point() +
  scale_x_log10()
```

The points appear to be spread out on the x-axis after the log transformation. This makes it easy to see that there isn't a correlation between `population` and life `expectancy`.

### The `color` and the `size` Aesthetics
```{r}
# Scatter plot comparing gdpPercap and lifeExp, with color representing continent
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color=continent)) +
  geom_point() +
  scale_x_log10()
```

This plot communicates a lot about differences in GDP and life expectancy among continents. For instance, the average life expectancy (and the GDP per capita to some extend) tends to be lowest for African countries.

We can also represent the `population` of the countries with `size` aesthetic.

```{r}
# Scatter plot comparing gdpPercap and lifeExp, with color representing continent
# Add the size aesthetic to represent a country's gdpPercap
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color=continent, size=pop, alpha=.5)) +
  geom_point() +
  scale_x_log10()
```

### Faceting

```{r}
# Scatter plot comparing gdpPercap and lifeExp, faceted by continent
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point(alpha=0.5) +
  scale_x_log10() +
  facet_wrap(~ continent)
```

### Visualizing Summarized Data

```{r}
# Summarize data within each year within each continent
by_year_continent <-  gapminder %>%
  group_by(year, continent) %>%
  summarize(meanLifeExp = mean(lifeExp),
            totalPop = sum(as.numeric(pop)))

# Plot the change in totalPop in each continent over time
ggplot(by_year_continent, aes(x = year, y = totalPop, color = continent)) +
  geom_point() +
  expand_limits(y = 0)
```

It looks like mean population across all continents (except Oceanica) is increasing over time.

```{r}
# Summarize the median GDP and median life expectancy per continent in 2007
by_continent_2007 <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp),
    medianGdpPercap = median(gdpPercap))

# Use a scatter plot to compare the median GDP and median life expectancy
ggplot(by_continent_2007, aes(x = medianGdpPercap, y = medianLifeExp, color = continent)) +
  geom_point() +
  expand_limits(y = 0)
```

### Line Plot

```{r}
# Summarize the median gdpPercap by year, then save it as by_year
by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap over time
ggplot(by_year, aes(x = year, y = medianGdpPercap)) +
  geom_line() +
  expand_limits(y = 0) +
  labs(title = "Change in Median GDP per Capita")
```

Median GDP per capita across countries has gone up over time. However the graph  below shows that the growth in median GDP per capita differ between continents.

```{r}
# Summarize the median gdpPercap by year & continent, save as by_year_continent
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap by continent over time
ggplot(by_year_continent, aes(x = year, y = medianGdpPercap, color = continent)) +
  geom_line() +
  expand_limits(y = 0) 
```

### Bar Plot

```{r}
# Summarize the median gdpPercap by year and continent in 2007
by_continent <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))


# Create a bar plot showing medianGdp by continent
ggplot(by_continent, aes(x = continent, y = medianGdpPercap, color = continent)) +
  geom_col()
```

```{r}
# Filter for observations in the Europe continent in 2007
europe_2007 <- gapminder %>%
  filter(year == 2007, continent == "Europe")
  
# Create a bar plot of gdpPercap by country
ggplot(europe_2007, aes(reorder(country, gdpPercap), gdpPercap)) +
  geom_col() + 
  coord_flip() + # horizontal bar plot
  ggtitle("Bar Plot of gdpPercap by Country in Europe in 2007")
```

### Histograms

```{r}
# Plot histogram of lifeExp
ggplot(gapminder_2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)
```

### Boxplots



```{r}
# Create a boxplot comparing lifeExp among continents
ggplot(gapminder_2007, aes(continent, lifeExp)) +
  geom_boxplot()
```


